\documentclass{beamer}

\usetheme{Madrid}
\usecolortheme{default}
\setbeamerfont{date}{size=\footnotesize}

\hypersetup{
  colorlinks=true,
  urlcolor=blue,
  linkcolor=lime,
  citecolor=magenta
}
\usepackage{xcolor}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, positioning, shapes.geometric}

\newcommand{\cverb}[1]{{\color{purple}{\texttt{#1}}}}

% --- Visual structure improvements (same as Talk 1) ---
\setbeamertemplate{section page}[simple]
\AtBeginSection{\frame{\sectionpage}}

\title[Maintaining a Large Formalization Project 2]{Maintaining a Large Formalization Project -- 2}
\subtitle{Part II: Automation \& AI-Specific Challenges}
\author{Damiano Testa}
\institute[Warwick]{University of Warwick}
\date{Tsinghua University, March 16-20, 2026}

\begin{document}

\begin{frame}
  \titlepage
\end{frame}

% =====================================================
\section{AI Failure Modes}
% =====================================================

% --- Slide: AI Errors ---
\begin{frame}{AI Errors}
In general, AI-generated definitions and statements tend to be

\medskip

\begin{itemize}
  \item more prone to misformalisations
  \medskip
  \item more inefficient
  \medskip
  \item harder to maintain
\end{itemize}

\medskip

than human-written ones.

\bigskip

Humans tend to spot quickly if a definition or statement diverges from its intended meaning.

\bigskip

AI agents tend to be less aware of context and expectations.

\bigskip

\begin{block}{Key Point}
Alignment with intended goals should be automatically enforced.
\end{block}

\end{frame}

% --- Slide: Exploits ---
\begin{frame}{Exploits}
AI might exploit Lean's flexible metaprogramming to
\begin{itemize}
  \item bypass kernel checking
  \smallskip
  \item tampering with build artifacts
  \smallskip
  \item sneakily introduce new \cverb{axiom}s
\end{itemize}

\bigskip

Partial protection: external checkers that replay the environment

\medskip

\begin{itemize}
  \item \href{https://github.com/fgdorais/lean4lean}{lean4lean}
  \smallskip
  \item \href{https://github.com/leanprover/leanchecker}{leanchecker}
  \smallskip
  \item \href{https://github.com/nanoda/}{nanoda}
  \smallskip
  \item \href{https://github.com/leanprover-community/lean-comparator}{comparator}
\end{itemize}

\medskip

See the \href{https://arena.lean-lang.org/}{Lean Kernel Arena} for more information.
\end{frame}

% --- Slide ---
\begin{frame}{Disclaimer on the Lean Kernel}
There is no known soundness issue in the Lean kernel.

\bigskip

But Lean metaprogramming can circumvent checks and create illusions of progress.

\bigskip

In a setup where there is an expectation that humans will not be able to check all the AI-generated code, this issue needs to be addressed.
\end{frame}

% =====================================================
\section{Reviewing AI Contributions}
% =====================================================

% --- Automatic Reviews ---
\begin{frame}{Automatic Reviews}
Guardrails for AI-authored PRs:

\medskip

\begin{itemize}
  \item check definitions against the blueprint
  \medskip
  \item ensure idiomatic constructs where possible
  \medskip
  \item adhere to project APIs
  \medskip
  \item automated lints against deviance from established guidelines
  \medskip
  \item expose all meta-programming to be really sure it is not a liability
\end{itemize}
\end{frame}

% --- Slide ---
\begin{frame}{AI Reviewing AI}
\begin{columns}[T,totalwidth=\textwidth]
  \begin{column}{0.6\textwidth}
Different AI agents excel at:

\medskip

\begin{itemize}
  \item understanding human arguments
  \smallskip
  \item formalisation
  \smallskip
  \item enforcing guidelines
  \smallskip
  \item suggesting improvements
\end{itemize}

\medskip

Harness these different agents to review AI-generated code.

\bigskip

For a project of this scope, human reviews will be at a premium.
  \end{column}
  \begin{column}{0.4\textwidth}
\begin{center}
\begin{tikzpicture}[
  >=Stealth,
  node distance=5.5mm,
  every node/.style={draw, rounded corners, align=center, inner sep=1.8pt, font=\small}
]
\node (formaliser) {AI Agent:\\Formaliser};
\node (lint) [below=of formaliser] {AI Agent:\\Lint Enforcer};
\node (api) [below=of lint] {AI Agent:\\API Auditor};
\node (refactor) [below=of api] {AI Agent:\\Refactorer};
\node (human) [below=8mm of refactor] {Human Review?};

\draw[->] (formaliser) -- (lint);
\draw[->] (lint) -- (api);
\draw[->] (api) -- (refactor);
\draw[->] (refactor) -- (human);
\end{tikzpicture}
\end{center}
  \end{column}
\end{columns}
\end{frame}

% --- Slide ---
\begin{frame}{External Re-Checking}

Require PRs to include provenance notes (prompt, model tag, seed/hash) for auditing.

\bigskip

Periodically:

\medskip

\begin{itemize}
  \item reassess reviewing criteria
  \medskip
  \item adapt to recent AI outputs
  \medskip
  \item look for common patterns, generalisations, further automation
\end{itemize}
\end{frame}

% =====================================================
\section{Automation Foundations}
% =====================================================

% --- Automation section ---
\begin{frame}{Automation}
As we discussed so far, fully AI-driven projects need checks at all levels.

\bigskip

After all, what good is a proof spanning several million lines of code, if \emph{anywhere} in the middle there is an undetected \cverb{sorry}?

\bigskip

This is a genuine and valid concern that must be addressed.

\bigskip

The ``default'' Lean settings prefer usability and flexibility over adversarial proof development.

\bigskip

This ``good faith'' principle can (and does) get violated by AI agents.
\end{frame}

% --- Slide ---
\begin{frame}{Generic Automation}
Automate:

\medskip

\begin{itemize}
  \item triage of PRs to the appropriate reviewers
  \smallskip
  \item task assignment, depending on contributors' strengths
  \smallskip
  \item iterate quickly on easy issues (typos, rewriting)
\end{itemize}

\medskip

Use priority queues matching blueprint dependency structure.

\bigskip

These suggestions apply mostly to human development.

\bigskip

However, I expect that some form of the above workflow applies to AI development as well.
\end{frame}

% =====================================================
\section{Lean-Specific Automation}
% =====================================================

% --- Slide ---
\begin{frame}{Lean-Specific Automation}
Figure out automatically reusable generalisations
\begin{itemize}
  \item identify common rewrites
  \medskip
  \item isolate subproofs and extract them to standalone lemmas
  \medskip
  \item explore new potential typeclasses to boost progress
\end{itemize}

\bigskip

Consider custom tactics — but AI-generated tactics need strict boundaries and tests.

\bigskip

Bad tactics are a huge liability.

\bigskip

A more advanced version of \cverb{group} would be useful.

\bigskip

As far as I know, there is no example of any serious tactic having ever been written autonomously by an AI.
\end{frame}

% --- Slide ---
\begin{frame}{Abstraction and Predictability}
When formalising, certain auxiliary concepts emerge.

\bigskip

These are often typeclasses or definitions that are not explicit in informal mathematics.

\bigskip

AIs should also have mechanisms for discovering such abstractions.

\bigskip

And they should also retain them in later iterations!

\bigskip

Developing new concepts benefits from predictability:

\smallskip

\begin{itemize}
  \item general definitions
  \smallskip
  \item abstraction via typeclasses
  \smallskip
  \item naming conventions
  \smallskip
  \item careful documentation
\end{itemize}
\end{frame}

% =====================================================
\section{Environment Shaping}
% =====================================================

% --- Slide ---
\begin{frame}{Helping Automation}
Some ``discovery'' tactics such as
\begin{itemize}
  \item \cverb{try?}
  \item \cverb{exact?}
  \item \cverb{apply?}
  \item \cverb{rw?}
\end{itemize}

are more performant with smaller environments.

\bigskip

Environment reduction can be achieved enforcing

\medskip

\begin{itemize}
  \item consistent module structure
  \smallskip
  \item semantic grouping of lemmas
  \smallskip
  \item predictable API
  \smallskip
  \item minimised \cverb{import}s
\end{itemize}
\end{frame}

% --- Slide ---
\begin{frame}{Helping Automation}

In a similar spirit, encouraging small, well-scoped files reduces conflicts.

\bigskip

Adhering to naming conventions helps formalisers use API lemmas.

\bigskip

Consistent module docs and docstrings certainly help humans and may also aid AI agents.

\bigskip

Extend dynamically the scope of \cverb{simp}, \cverb{aesop}, \cverb{grind}, by training agents to tag lemmas appropriately.
\end{frame}

% --- Slide ---
\begin{frame}{Developing Custom Tactics}
Where recurring patterns appear:

\medskip

\begin{itemize}
  \item create domain-specific \cverb{simp}-sets
  \medskip
  \item expand \cverb{aesop} and \cverb{grind} when appropriate
  \medskip
  \item create domain-specific tactics
\end{itemize}

\bigskip

Two or three well-written tactics can save hundreds of lines.
\end{frame}

% =====================================================
\section{Computation \& Blueprinting}
% =====================================================

% --- Slide ---
\begin{frame}{Bird's Eye View for Integrated Development}
\begin{columns}[T,totalwidth=\textwidth]
  \begin{column}{0.45\textwidth}
Provide:

\medskip

\begin{itemize}
  \item dashboards
  \medskip
  \item dependency graphs
  \medskip
  \item coverage of key results
  \medskip
  \item ``blocked by'' relationships
\end{itemize}
  \end{column}
  \begin{column}{0.6\textwidth}
\begin{tikzpicture}[
  >=Stealth,
  node distance=8mm,
  every node/.style={draw, rounded corners, align=center, inner sep=2pt, font=\scriptsize}
]
\node (A) {Lemma A};
\node (B) [below left=8mm and 10mm of A] {Lemma B};
\node (C) [below right=8mm and 10mm of A] {Lemma C};
\node (D) [below=20mm of A] {Lemma D};
\node (E) [below=6mm of D] {Theorem E};

\draw[->] (A) -- (B);
\draw[->] (A) -- (C);
\draw[->] (B) -- (D);
\draw[->] (C) -- (D);
\draw[->] (D) -- (E);
\end{tikzpicture}
  \end{column}
\end{columns}

\bigskip

Helps parallelisation and spotting blockers.

\bigskip

Generally, any form of visual feedback is very useful for humans.

\bigskip

(Possibly not for AIs, but it is a good way to track progress.)
\end{frame}

% --- Slide ---
\begin{frame}{Proofs by Computation}
\begin{columns}[T,totalwidth=\textwidth]
  \begin{column}{0.58\textwidth}
Parts of the classification require explicit computations:

\medskip

\begin{itemize}
  \item finite case splits
  \medskip
  \item numeric inequalities
  \medskip
  \item small search spaces
  \medskip
  \item character computations
\end{itemize}

\bigskip

Mathlib has limited infrastructure for this currently.

\bigskip

The classification of finite simple groups will benefit from trusted computations.
  \end{column}
  \begin{column}{0.42\textwidth}
\begin{center}
\begin{tikzpicture}[
  >=Stealth,
  node distance=7mm,
  every node/.style={draw, rounded corners, align=center, inner sep=1.6pt, font=\scriptsize}
]
\node (input) {Input Data};
\node (ext) [below=of input] {External Computation\\(Macaulay2/GAP/Python)\\\textit{(untrusted)}};
\node (ver) [below=of ext] {Verification Layer\\(Lean checks/bridges)};
\node (lean) [below=of ver] {Lean Theorem};

\draw[->] (input) -- (ext);
\draw[->] (ext) -- (ver);
\draw[->] (ver) -- (lean);
\end{tikzpicture}
\end{center}
  \end{column}
\end{columns}
\end{frame}

% --- Slide ---
\begin{frame}{Mathematics: Blueprint}
The blueprint is a bridge between the original proof and the final target.

\bigskip

It should probably be part of

\medskip

\begin{itemize}
  \item the training data and
  \smallskip
  \item the final formalisation.
\end{itemize}

\bigskip

Its history is a good indicator of successes and failures of the formalisation.
\end{frame}

% --- Slide ---
\begin{frame}{Training on Past Versions}
Training AI agents to \emph{write} good blueprints is probably a good intermediate target.

\bigskip

Keep the blueprint up-to-date and treat it as:

\medskip

\begin{itemize}
  \item a source of truth
  \smallskip
  \item an intermediate representation mapping informal → formal goals
\end{itemize}

For AI-driven projects, past versions could still be used as training data.

\bigskip

Re-using code that worked well is a way of consolidating good practices.

\bigskip

Future iterations can then learn good practices and improve further.
\end{frame}

% =====================================================
\section{Reducing Human Burden}
% =====================================================

% --- Slide ---
\begin{frame}{Support to Reduce Human Input}
Linters can flag deviations from good practices, both
\begin{itemize}
  \item locally (typically syntax linters) and
  \item globally (typically environment linters).
\end{itemize}

\bigskip

CI can integrate automatically the linter suggestions: it is important that the linter suggestions are actionable!

\bigskip

The ``informalisation'' process that goes from a maths paper intended for humans to a text that is adapted to formalisation is very laborious.

\bigskip

Dedicated agents that provide details, fill in gaps, suggest intermediate goals, state results in a ``formalisation friendly'' are immensely valuable.

\bigskip

Human input and vetting will still be needed, but only for deeper issues.
\end{frame}

% =====================================================
\section{AI Pipeline Architecture}
% =====================================================

% --- Final slide ---
\begin{frame}{Auto-Formalisation Skeleton Layout}
\begin{columns}[T,totalwidth=\textwidth]
  \begin{column}{0.58\textwidth}
Possible layered AI approach:

\medskip

\begin{itemize}
  \item Expand blueprint
  \smallskip
  \item Split argument into informal lemmas
  \smallskip
  \item Write formal statements with \cverb{sorry}s
  \smallskip
  \item Verify alignment with blueprint
  \smallskip
  \item Iterate until agreement
  \smallskip
  \item Fill in \cverb{sorry}s
  \smallskip
  \item Use agents to try to disprove claims
  \smallskip
  \item Golfing agents simplify proofs
  \smallskip
  \item Generalisation agents detect patterns
  \smallskip
  \item Refactor based on insights
\end{itemize}
  \end{column}
  \begin{column}{0.42\textwidth}
\begin{center}
\begin{tikzpicture}[
  >=Stealth,
  node distance=4mm,
  every node/.style={draw, rounded corners, align=center, inner sep=1.6pt, font=\scriptsize, text width=0.9\linewidth}
]
\node (b) {Expand Blueprint};
\node (s) [below=of b] {Split Into Informal Lemmas};
\node (g) [below=of s] {Generate Lean Code with \cverb{sorry}s};
\node (c) [below=of g] {Check Blueprint Alignment};
\node (f) [below=of c] {Fill In \cverb{sorry}s};
\node (a) [below=of f] {Look for Counterexamples};
\node (golf) [below=of a] {Golfing Agents Simplify};
\node (gen) [below=of golf] {Generalisation Agents};
\node (r) [below=of gen] {Refactor Based on Insights};

\draw[->] (b) -- (s);
\draw[->] (s) -- (g);
\draw[->] (g) -- (c);
\draw[->] (c) -- (f);
\draw[->] (f) -- (a);
\draw[->] (a) -- (golf);
\draw[->] (golf) -- (gen);
\draw[->] (gen) -- (r);
\end{tikzpicture}
\end{center}
  \end{column}
\end{columns}
\end{frame}

% =====================================================
\section{Short Proofs and Lemmas}
% =====================================================

\begin{frame}{Short Proofs}
Short proofs improve:
\medskip
\begin{itemize}
  \item readability
  \item review time
  \item maintainability
  \item conceptual clarity
\end{itemize}
\medskip
Long monolithic proofs accumulate accidental complexity.
\end{frame}

\begin{frame}{Atomic Lemmas}
Breaking arguments into lemmas provides:
\medskip
\begin{itemize}
  \item reusable API-level statements
  \item recurring subgoals that guide structure
  \item clearer opportunities for abstraction
  \item smaller, local proof obligations
\end{itemize}
This is essential for long-term maintenance.
\end{frame}

\begin{frame}{Walls of \cverb{have}}
Forward reasoning via many \cverb{have}s produces:
\medskip
\begin{itemize}
  \item opaque clutter
  \item loss of goal structure
  \item brittle, linear dependency chains
  \item hidden lemma candidates
\end{itemize}
Backward reasoning exposes structure; walls of \cverb{have} conceal it.
\end{frame}

\begin{frame}{Backward Reasoning}
Lean's tactic framework is built around:
\medskip
\begin{itemize}
  \item goal transformation (\cverb{apply}, \cverb{refine})
  \item structural decomposition (\cverb{intro}, \cverb{cases})
  \item clean subgoal generation
\end{itemize}
This naturally yields maintainable, idiomatic proofs.
\end{frame}

\begin{frame}{AI Succeeds Because Mathlib Exists}
Modern AI proof agents succeed partly because Mathlib:
\medskip
\begin{itemize}
  \item provides thousands of general, well-named lemmas
  \item offers predictable and curated APIs
  \item encodes human-designed abstractions AI can retrieve
\end{itemize}
AI can take a few steps ahead only because humans built strong foundations.
\end{frame}

\begin{frame}{Current AI Proofs Do Not Scale}
AI-generated proofs tend to be:
\medskip
\begin{itemize}
  \item verbose
  \item narrowly optimised
  \item hard to maintain
  \item hard to reuse
\end{itemize}
Such code cannot scale to CFSG-level formalisation.
\end{frame}

\begin{frame}{Mathlib vs.\ CFSG: Scale}
Approximate size comparison:
\medskip
\begin{itemize}
  \item Mathlib corresponds to $\sim$5k--10k pages of informal mathematics.
  \item The CFSG spans $\sim$10k--20k pages.
\end{itemize}

\bigskip

Scaling requires Mathlib-level quality throughout:

\medskip

\begin{itemize}
  \item modular lemmas
  \item predictable APIs
  \item short proofs
\end{itemize}
\end{frame}

\begin{frame}{On Maintainability (after Avigad)}
Public essays and talks by Avigad emphasise:

\medskip

\begin{itemize}
  \item AI proofs are often verbose and narrowly targeted
  \item maintainability must remain a primary goal
  \item incentives should promote refactoring, not credit accumulation
\end{itemize}

\bigskip

This matches concerns raised widely regarding AI-driven formalisation.
\end{frame}

% --- Thanks! ---
\begin{frame}
\begin{center}
  {\huge Thank You!}\\[30pt]
  {\huge Questions?}
\end{center}
\end{frame}

\end{document}
